---
title: "LEGUME_EXUDATE_GC-MS&ENRICHMENT FIGURES-EXUDATA"
author: "AA"
date: "2026-09-10"
output: html_document
editor_options: 
  chunk_output_type: console
---

# GC-MS LEGUME ROOT EXUDATE ANALYSIS

```{r}
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(mixOmics)
library(DescTools)
library(grid)
```

# 2. Import and prepare data

```{r}
df <- read.csv("Ex_data2.csv", check.names = FALSE)

metabolite_info <- data.frame(
  Metabolite = colnames(df)[-1],
  Main_class = as.character(df[1, -1]),
  Sub_class = as.character(df[2, -1])
)

metabolite_info$Main_class[
  is.na(metabolite_info$Main_class) |
    metabolite_info$Main_class == "NA"
] <- "Unclassified"

metabolite_info$Sub_class[
  is.na(metabolite_info$Sub_class) |
    metabolite_info$Sub_class == "NA"
] <- "Unclassified"

abundance <- df[-c(1, 2), ]
rownames(abundance) <- abundance[, 1]
abundance <- abundance[, -1]
abundance[] <- lapply(abundance, as.numeric)

sample_info <- data.frame(
  Sample = rownames(abundance),
  Group = c(
    "Bean", "Bean", "Bean",
    "Cowpea", "Cowpea", "Cowpea",
    "Pea", "Pea", "Pea"
  )
)

sample_info$Variety <- dplyr::recode(
  sample_info$Group,
  Bean = "Anasazi",
  Cowpea = "California5",
  Pea = "Vail"
)
```

# 3. Normalization and log2 transformation

```{r}
sample_medians <- apply(
  abundance,
  1,
  median,
  na.rm = TRUE
)

global_median <- median(
  sample_medians,
  na.rm = TRUE
)

abundance_norm <- sweep(
  abundance,
  1,
  sample_medians,
  "/"
) * global_median

abundance_log2 <- log2(
  abundance_norm + 1
)
```

# 4. G-test

```{r}
g_test_results <- lapply(
  colnames(abundance),
  function(met) {

    detected <- abundance[[met]] > 0

    tab <- table(
      Group = sample_info$Group,
      Detected = detected
    )

    bean_detected <- sum(
      detected[sample_info$Group == "Bean"]
    )

    cowpea_detected <- sum(
      detected[sample_info$Group == "Cowpea"]
    )

    pea_detected <- sum(
      detected[sample_info$Group == "Pea"]
    )

    highest_detected <- c(
      Bean = bean_detected,
      Cowpea = cowpea_detected,
      Pea = pea_detected
    )

    highest_group <- names(
      which.max(highest_detected)
    )

    if (ncol(tab) < 2 || nrow(tab) < 2) {

      return(
        data.frame(
          Metabolite = met,
          Bean_detected = bean_detected,
          Cowpea_detected = cowpea_detected,
          Pea_detected = pea_detected,
          Highest_detected = highest_group,
          G_statistic = NA,
          p_value = NA
        )
      )
    }

    gt <- tryCatch(
      DescTools::GTest(
        tab,
        correct = "williams"
      ),
      error = function(e) NULL
    )

    if (is.null(gt)) {

      return(
        data.frame(
          Metabolite = met,
          Bean_detected = bean_detected,
          Cowpea_detected = cowpea_detected,
          Pea_detected = pea_detected,
          Highest_detected = highest_group,
          G_statistic = NA,
          p_value = NA
        )
      )
    }

    data.frame(
      Metabolite = met,
      Bean_detected = bean_detected,
      Cowpea_detected = cowpea_detected,
      Pea_detected = pea_detected,
      Highest_detected = highest_group,
      G_statistic = unname(gt$statistic),
      p_value = gt$p.value
    )
  }
)

g_test_results <- do.call(
  rbind,
  g_test_results
)

g_test_results$FDR <- p.adjust(
  g_test_results$p_value,
  method = "fdr"
)

g_test_results <- g_test_results %>%
  left_join(
    metabolite_info,
    by = "Metabolite"
  ) %>%
  arrange(FDR)
```

# 5. PLS-DA

```{r}

# 5. PLS-DA

# Remove metabolites with zero variance
metabolite_sd <- apply(
  abundance_log2,
  2,
  sd,
  na.rm = TRUE
)

abundance_pls <- abundance_log2[
  ,
  metabolite_sd > 0,
  drop = FALSE
]

# Define plant groups
Y <- factor( sample_info$Variety,
  levels = c( "Anasazi", "California5",
    "Vail")
)
# Run PLS-DA
plsda_model <- mixOmics::plsda(
  X = abundance_pls,
  Y = Y,
  ncomp = 2
)

# Extract PLS-DA coordinates
plsda_df <- data.frame(
  Sample = rownames(abundance_pls),
  Group = Y,
  Comp1 = plsda_model$variates$X[, 1],
  Comp2 = plsda_model$variates$X[, 2]
)

# Variance explained
plsda_var <- plsda_model$prop_expl_var$X * 100

comp1_var <- round(
  plsda_var[1],
  1
)
comp2_var <- round(
  plsda_var[2],
  1
)

# Colors
group_colors <- c(
  Anasazi = "#8B0000",
  California5 = "#B8860B",
  Vail = "#00008B"
)

# PLS-DA figure
plsda_plot <- ggplot(  plsda_df,
  aes(  x = Comp1, y = Comp2,
    color = Group, label = Sample )) +
  geom_point(
    size = 4
  ) +
  geom_text_repel(
    size = 4,
    show.legend = FALSE
  ) +
  scale_color_manual(
    values = group_colors
  ) +
  theme_classic(
    base_size = 14
  ) +
  labs(
    title = "PLS-DA of Legume Root Exudate Metabolites",
    x = paste0(
      "Component 1 (",
      comp1_var,
      "%)"
    ),
    y = paste0(
      "Component 2 (",
      comp2_var,
      "%)"
    ),
    color = "Legume"
  ) +
  theme(
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    ),
    axis.title = element_text(
      face = "bold"
    ),
    axis.text = element_text(
      face = "bold"
    ),
    legend.title = element_text(
      face = "bold"
    ),
    legend.text = element_text(
      face = "bold"
    )
  )

print(plsda_plot)
```

# 6. VIP scores and group means

```{r}
group_means <- data.frame(
  Metabolite = colnames(abundance_log2),
  Bean_mean = colMeans(
    abundance_log2[
      sample_info$Group == "Bean",
      ,
      drop = FALSE
    ],
    na.rm = TRUE
  ), Cowpea_mean = colMeans( abundance_log2[
      sample_info$Group == "Cowpea",
      ,  drop = FALSE ],
    na.rm = TRUE ),
  Pea_mean = colMeans( abundance_log2[ sample_info$Group == "Pea",
      , drop = FALSE ], na.rm = TRUE ))

vip_scores <- mixOmics::vip(
  plsda_model
)

vip_df <- data.frame(
  Metabolite = colnames(abundance_pls),
  VIP = vip_scores[, 1]
)

vip_df <- vip_df %>%
  left_join(
    metabolite_info,
    by = "Metabolite"
  ) %>%
  left_join(
    group_means,
    by = "Metabolite"
  ) %>%
  mutate(
    Highest_in = case_when(
      Bean_mean > Cowpea_mean &
        Bean_mean > Pea_mean ~ "Bean",
      Cowpea_mean > Bean_mean &
        Cowpea_mean > Pea_mean ~ "Cowpea",
      Pea_mean > Bean_mean &
        Pea_mean > Cowpea_mean ~ "Pea",
      TRUE ~ "Tie"
    )
  ) %>%
  arrange(
    desc(VIP)
  )
```

# 7. One-way ANOVA and FDR

```{r}
anova_results <- data.frame(
  Metabolite = colnames(abundance_log2),
  pvalue = NA_real_
)

for (i in seq_along(colnames(abundance_log2))) {

  met <- colnames(abundance_log2)[i]

  df_anova <- data.frame(
    value = abundance_log2[, met],
    Group = sample_info$Group
  )
# Remove missing values
  df_anova <- df_anova[
    complete.cases(df_anova),
  ]
# Run ANOVA
  model <- aov(
    value ~ Group,
    data = df_anova
  )
# Save ANOVA p-value
  anova_results$pvalue[i] <-
    summary(model)[[1]][["Pr(>F)"]][1]
}
# FDR correction
anova_results$FDR <- p.adjust(
  anova_results$pvalue,
  method = "fdr"
)
anova_results <- anova_results %>%
  left_join(
    metabolite_info,
    by = "Metabolite"
  ) %>%
  left_join(
    group_means,
    by = "Metabolite"
  ) %>%
  mutate(
    Highest_in = case_when(
 Bean_mean > Cowpea_mean &
        Bean_mean > Pea_mean ~ "Bean",

 Cowpea_mean > Bean_mean &
        Cowpea_mean > Pea_mean ~ "Cowpea",

 Pea_mean > Bean_mean &
        Pea_mean > Cowpea_mean ~ "Pea",
      TRUE ~ "Tie"
    )
  ) %>%
  arrange(FDR)
sig_anova <- anova_results %>%
  filter(
    !is.na(FDR),
    FDR < 0.05
  )
```

# 8. Heatmap of ANOVA-significant metabolites

```{r}
sig_metabolites <- sig_anova$Metabolite

heat_group <- data.frame(
  Anasazi = colMeans(
    abundance_log2[
      sample_info$Group == "Bean",
      sig_metabolites,
      drop = FALSE
    ],
    na.rm = TRUE
  ),
 California5 = colMeans(
    abundance_log2[
      sample_info$Group == "Cowpea",
      sig_metabolites,
      drop = FALSE
    ],
    na.rm = TRUE
  ),
  Vail = colMeans(
    abundance_log2[
      sample_info$Group == "Pea",
      sig_metabolites,
      drop = FALSE
    ],
    na.rm = TRUE
  )
)
rownames(
  heat_group
) <- sig_metabolites

heat_group_scaled <- t(
  scale(
    t(
      as.matrix(
        heat_group ) )))

heat_colors <- colorRampPalette(
  c(
    "#08306B",
    "#F7F7F7",
    "#99000D"
  )
)(100)

category_map <- c(
  "phthalic acid" = "Phenolic/aromatic",
  "3-phenyllactic acid" = "Phenolic/aromatic",
  "1-(2-hydroxyethyl)-2,2,6,6-tetramethyl-4-piperidinol*" = "Other",
  "xylo-pentodialdose*" = "Sugar/organic acid",
  "D-xylitol" = "Sugar/organic acid",
  "4-hydroxycinnamic acid (p-coumaric acid)" = "Phenolic/aromatic",
  "caffeic acid" = "Phenolic/aromatic",
  "4-hydroxypent-3-enoic acid*" = "Sugar/organic acid",
  "isophthalic acid*" = "Phenolic/aromatic",
  "bis(2-hydroxyethyl) phthalate*" = "Phenolic/aromatic",
  "2-(hydroxymethyl)-2-methylpropane-1,3-diol*" = "Other",
  "L-pyroglutamic acid" = "Amino acid/defense",
  "L-valine" = "Amino acid/defense",
  "L-phenylalanine" = "Amino acid/defense",
  "variabilin*" = "Amino acid/defense",
  "willardiine" = "Amino acid/defense",
  "L-homoserine" = "Amino acid/defense",
  "flavone base+2O+2MeO" = "Phenolic/aromatic",
  "L-glutamic acid" = "Amino acid/defense",
  "pisatin*" = "Amino acid/defense",
  "L-threonine" = "Amino acid/defense",
  "uracil" = "Amino acid/defense",
  "2,4-dihydroxybenzoic acid*" = "Phenolic/aromatic",
  "D-ribitol" = "Sugar/organic acid",
  "erythritol/threitol" = "Sugar/organic acid",
  "2,3,4-trihydroxybutyric acid *" = "Sugar/organic acid",
  "glycolic acid" = "Sugar/organic acid",
  "D-xylulose" = "Sugar/organic acid",
  "nonanoic acid" = "Sugar/organic acid"
)
row_annotations <- data.frame(
  Category = category_map[
    rownames(
      heat_group_scaled
    )
  ]
)
rownames(
  row_annotations
) <- rownames(
  heat_group_scaled
)
row_annotations$Category[
  is.na(row_annotations$Category)
] <- "Other"
row_colors <- list(
  Category = c(
    "Sugar/organic acid" = "#B8860B",
    "Phenolic/aromatic" = "#8B0000",
    "Amino acid/defense" = "#00008B",
    "Other" = "#808080"
  )
)
heatmap_plot <- pheatmap(
  heat_group_scaled,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  color = heat_colors,
  annotation_row = row_annotations,
  annotation_colors = row_colors,
  fontsize_row = 10,
  fontsize_col = 14,
  border_color = NA,
  main = "ANOVA-Significant Metabolites",
  legend = TRUE,
  silent = TRUE
)
grid.newpage()
grid.draw(
  heatmap_plot$gtable
)
```

# 9. Master table: ANOVA + VIP + G-test

```{r}
anova_merge <- anova_results %>%
  dplyr::rename(
    ANOVA_pvalue = pvalue,
    ANOVA_FDR = FDR,
    ANOVA_Highest = Highest_in
  )
vip_merge <- vip_df %>%
  dplyr::select(
    Metabolite,
    VIP,
    Highest_in
  ) %>%
  dplyr::rename(
    VIP_Highest = Highest_in
  )
gtest_merge <- g_test_results %>%
  dplyr::select(
    Metabolite,
    G_statistic,
    p_value,
    FDR,
    Highest_detected
  ) %>%
  dplyr::rename(
    GTest_pvalue = p_value,
    GTest_FDR = FDR,
    GTest_Highest = Highest_detected
  )

master_results <- anova_merge %>%
  left_join(
    vip_merge,
    by = "Metabolite"
  ) %>%
  left_join(
    gtest_merge,
    by = "Metabolite"
  ) %>%
  mutate(
    GTest_Highest = ifelse(
      is.na(G_statistic),
      NA_character_,
      GTest_Highest
    ),
    VIP_selected = ifelse(
      !is.na(VIP) & VIP >= 1,
      "Yes",
      "No"
    ),
    ANOVA_significant = ifelse(
      !is.na(ANOVA_FDR) &
        ANOVA_FDR < 0.05,
      "Yes",
      "No"
    ),
    GTest_significant = ifelse(
      !is.na(GTest_FDR) &
        GTest_FDR < 0.05,
      "Yes",
      "No"
    )
  ) %>%
  arrange(
    ANOVA_FDR,
    desc(VIP),
    GTest_FDR
  )

View(master_results)
```

# 10. Prepare enrichment table
```{r}
master <- master_results %>%
  dplyr::transmute(
    Metabolite = as.character(Metabolite),
    Anasazi = as.numeric(Bean_mean),
    California5 = as.numeric(Cowpea_mean),
    Vail = as.numeric(Pea_mean),
    pvalue = as.numeric(ANOVA_pvalue),
    FDR = as.numeric(ANOVA_FDR),
    Main_class = as.character(Main_class),
    Sub_class = as.character(Sub_class),
    HighestPlant = case_when(
      ANOVA_Highest == "Bean" ~ "Anasazi",
      ANOVA_Highest == "Cowpea" ~ "California5",
      ANOVA_Highest == "Pea" ~ "Vail",
      TRUE ~ NA_character_
    )
  ) %>%
  mutate(
    Anasazi_enrichment =
      Anasazi -
      ((California5 + Vail) / 2),
    California5_enrichment =
      California5 -
      ((Anasazi + Vail) / 2),
    Vail_enrichment =
      Vail -
      ((Anasazi + California5) / 2)
  )
```

# 11. Chemical categories for enrichment figure

```{r}
master <- master %>%
  mutate(
    Classification_text = paste(
      Main_class,
      Sub_class,
      Metabolite
    ),
    Chemical_group = case_when(
      grepl(
        "phenol|phenolic|aromatic|benzen|phenyl|flavonoid|cinnamic|coumaric|caffeic|ferulic|sinapic|gallic|phthal",
        Classification_text,
        ignore.case = TRUE
      ) ~ "Phenolic/aromatic",
      grepl(
        "amino|peptide|nitrogen|defense|glutam|aspart|alanine|valine|leucine|isoleucine|glycine|serine|threonine|lysine|arginine|histidine",
        Classification_text,
        ignore.case = TRUE
      ) ~ "Amino acid/defense",
      grepl(
        "organic acid|carboxylic|dicarboxylic|hydroxy acid|glycolic acid|nonanoic acid|trihydroxybutyric acid|malonic|succinic|citric|malic|fumaric|aconitic|quinic|pent.*enoic",
        Classification_text,
        ignore.case = TRUE
      ) ~ "Organic acid",
      grepl(
        "sugar|carbohydrate|saccharide|sugar alcohol|xylitol|xylulose|glucose|fructose|sucrose|trehalose|ribose|xylose|fucose|sorbitol|mannitol",
        Classification_text,
        ignore.case = TRUE
      ) ~ "Sugar/sugar alcohol",
      TRUE ~ "Other/Unclassified"
    ),
    Category = case_when(
      Chemical_group %in% c(
        "Sugar/sugar alcohol",
        "Organic acid"
      ) ~ "Sugar/organic acid",
      TRUE ~ Chemical_group
    )
  )
```

# 12. Enrichment figure settings
```{r}
category_colors <- c(
  "Phenolic/aromatic" = "#8B0000",
  "Amino acid/defense" = "#1F4E79",
  "Sugar/organic acid" = "#C99700",
  "Other/Unclassified" = "#6E6E6E"
)

all_enrichment_values <- c(
  master$Anasazi_enrichment[
    master$HighestPlant == "Anasazi" &
      master$FDR < 0.05
  ],
  master$California5_enrichment[
    master$HighestPlant == "California5" &
      master$FDR < 0.05
  ],
  master$Vail_enrichment[
    master$HighestPlant == "Vail" &
      master$FDR < 0.05
  ]
)

all_enrichment_values <- all_enrichment_values[
  is.finite(all_enrichment_values) &
    all_enrichment_values > 0
]

common_xmax <- ceiling(
  max(
    all_enrichment_values,
    na.rm = TRUE
  ) * 1.08
)

# 13. Enrichment figure function

plot_plant_enrichment <- function(
    plant,
    top_n = 15
) {

  enrichment_column <- paste0(
    plant,
    "_enrichment"
  )

  df_plot <- master %>%
    filter(
      HighestPlant == plant
    ) %>%
    mutate(
      Enrichment =
        .data[[enrichment_column]]
    ) %>%
    filter(
      !is.na(Enrichment),
      is.finite(Enrichment),
      Enrichment > 0,
      !is.na(FDR),
      FDR < 0.05
    ) %>%
    arrange(
      FDR,
      desc(Enrichment)
    ) %>%
    slice_head(
      n = top_n
    ) %>%
    mutate(
      Significance = case_when(
        FDR < 0.001 ~ "***",
        FDR < 0.01 ~ "**",
        FDR < 0.05 ~ "*",
        TRUE ~ ""
      ),
      PlotLabel = paste0(
        Metabolite,
        " ",
        Significance
      )
    )

  df_plot$Category <- factor(
    df_plot$Category,
    levels = c(
      "Phenolic/aromatic",
      "Amino acid/defense",
      "Sugar/organic acid",
      "Other/Unclassified"
    )
  )

  df_plot <- df_plot %>%
    arrange(
      Category,
      desc(Enrichment)
    )

  df_plot$PlotLabel <- factor(
    df_plot$PlotLabel,
    levels = rev(
      df_plot$PlotLabel
    )
  )

  display_name <- ifelse(
    plant == "California5",
    "California-5",
    plant
  )

  ggplot(
    df_plot,
    aes(
      x = Enrichment,
      y = PlotLabel,
      fill = Category
    )
  ) +
    geom_col(
      width = 0.70
    ) +
    facet_grid(
      rows = vars(Category),
      scales = "free_y",
      space = "free_y",
      switch = "y"
    ) +
    scale_fill_manual(
      values = category_colors,
      drop = FALSE
    ) +
    scale_x_continuous(
      limits = c(
        0,
        common_xmax
      ),
      expand = expansion(
        mult = c(
          0,
          0.03
        )
      )
    ) +
    labs(
      title = display_name,
      subtitle = "ANOVA-significant metabolites (FDR < 0.05)",
      x = "Enrichment relative to the mean\nof the other two plants (log2)",
      y = NULL
    ) +
    theme_classic(
      base_size = 14
    ) +
    theme(
      plot.title = element_text(
        face = "bold",
        size = 18,
        hjust = 0.5
      ),
      plot.subtitle = element_text(
        face = "bold",
        size = 11,
        hjust = 0.5,
        margin = margin(
          b = 10
        )
      ),
      axis.title.x = element_text(
        face = "bold",
        size = 13,
        margin = margin(
          t = 15
        )
      ),
      axis.text.x = element_text(
        face = "bold",
        size = 11
      ),
      axis.text.y = element_text(
        face = "bold",
        size = 10.5
      ),
      strip.text.y.left = element_text(
        angle = 0,
        face = "bold",
        size = 10.5
      ),
      strip.background = element_rect(
        fill = "grey95",
        color = "grey40",
        linewidth = 0.6
      ),
      legend.position = "none",
      panel.spacing.y = grid::unit(
        0.45,
        "lines"
      )
    )
}
```

# 14. Create enrichment figures

```{r}
p_Anasazi <- plot_plant_enrichment(
  "Anasazi",
  top_n = 15
)

p_California5 <- plot_plant_enrichment(
  "California5",
  top_n = 15
)

p_Vail <- plot_plant_enrichment(
  "Vail",
  top_n = 15
)

print(p_Anasazi)
print(p_California5)
print(p_Vail)
```

# 15. Save final outputs

```{r}
dir.create(
  "Fig5",
  showWarnings = FALSE,
  recursive = TRUE
)
ggsave(
  "Fig5/PLSDA_Legume_Root_Exudates.png",
  plsda_plot,
  width = 7,
  height = 5,
  dpi = 600,
  bg = "white"
)
png(
  "Fig5/ANOVA_significant_heatmap.png",
  width = 2400,
  height = 3000,
  res = 300
)
grid.draw(
  heatmap_plot$gtable
)

dev.off()

ggsave(
  "Fig5/Anasazi_enrichment.png",
  p_Anasazi,
  width = 10,
  height = 8,
  dpi = 600,
  bg = "white"
)

ggsave(
  "Fig5/California5_enrichment.png",
  p_California5,
  width = 10,
  height = 8,
  dpi = 600,
  bg = "white"
)

ggsave(
  "Fig5/Vail_enrichment.png",
  p_Vail,
  width = 10,
  height = 8,
  dpi = 600,
  bg = "white"
)

write.csv(
  master_results,
  "Fig5/Master_ANOVA_VIP_GTest_Results.csv",
  row.names = FALSE
)

write.csv(
  anova_results,
  "Fig5/ANOVA_FDR_results.csv",
  row.names = FALSE
)

write.csv(
  vip_df,
  "Fig5/VIP_scores.csv",
  row.names = FALSE
)

write.csv(
  g_test_results,
  "Fig5/GTest_results.csv",
  row.names = FALSE
)
```

